STR: Seasonal-Trend decomposition using Regression

### 2021-03-29 str_ijds.tex
Reverted to original ordering of authors
Nearly ready to resubmit

\author{Rob J~Hyndman, Alexander~Dokumentov}
\addresses{\textbf{Rob J Hyndman}\newline
Monash University\newline Australia
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\newline Corresponding author\\[1cm]
\textbf{Alexander Dokumentov}\newline
Growing Data\newline Australia
\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Hyndman, Dokumentov: \@date}
\makeatother

\graphicspath{{figs/}}
\def\addlinespace{}
\usepackage{amsmath}
\setlength{\arraycolsep}{3pt}
\setcounter{MaxMatrixCols}{20}
\newcommand{\argmin}{\operatorname{arg\,min}}
\newcommand{\sn}{\kappa}
\renewcommand{\vec}{\operatorname{vec}}
\newcommand{\diag}{\operatorname{diag}}
\renewcommand{\mod}{\operatorname{mod}}
\usepackage[section]{placeins}

@@

The first attempts to decompose time series into various components can be dated to the 19th century, when \textcite{Poynting1884} proposed price averaging as a tool for eliminating trend and seasonal fluctuations. Later, his approach was extended by \textcite{hooker1901suspension}, \textcite{spencer1904graduation} and \textcite{anderson1914nochmals}. Further research in this direction included that by \textcite{copeland1915statistical}, who was the first to attempt to extract the seasonal component, until \textcite{macaulay1930smooth} proposed a method which gradually became ``classical''. The work of Macaulay led to the Census II method \autocite{shiskin1957electronic} which became popular amongst national statistics offices. The Census II method has continued to evolve, with additional features including robustness, calendar effects, covariates, ARIMA extensions, and extensive diagnostics. Widely used versions of this approach have been X-11 \autocite{shishkin1967x}, X-11-ARIMA \autocite{dagum1988x11arima,ladiray2001seasonal}, X-12-ARIMA \autocite{findley1998new} and X-13-ARIMA-SEATS \autocite{findley2005some}.

@@

Several model-based methods for seasonal decomposition have been developed, including the TRAMO/SEATS procedure which was developed at the Bank of Spain \autocite{Gomez2001}, the TBATS model of \textcite{de2011forecasting}, and various structural time series model approaches \autocite{harvey1990forecasting,Commandeur2011}. One big advantage of using a model for seasonal decomposition and adjustment is that it provides a natural way to compute confidence and prediction intervals.

@@

@@

@@

@@

We obtained similar results using other decomposition methods applied to the same data including STL \autocite{cleveland1990stl} and X-13-ARIMA-SEATS \autocite{findley2005some}.

@@

This data set has two seasonal patterns. The first pattern is a weekly seasonal pattern (\(m_1=48\times7=336\)) that represents the specific demand features that are attributable to a particular day of the week. The second pattern is a daily seasonal pattern (\(m_2=48\)) with the topology in Figure \ref{fig:topology2_pdf}, which allows the model to distinguish between working days and holidays/weekends, and to make transitions between them. The pattern reflects the tendency to have a higher electricity demand during standard working hours and a lower demand at night. It also reflects the tendency to have different demand patterns on working days and holidays/weekends. A longer series would also have an annual seasonal pattern, but with only 115 days, this cannot be distinguished from the trend component.

@@

\begin{table}

\caption{\label{tab:outlierstable}The ten residuals that are largest in absolute value after an STR decomposition.}
\centering
\begin{tabular}[t]{rlcr}
\toprule
Date & Day of week & Time period & Residual\\
\midrule
3 February 2000 & Thursday & 14:30 -- 15:00 & 687.3\\
3 February 2000 & Thursday & 15:00 -- 15:30 & 871.3\\
3 February 2000 & Thursday & 15:30 -- 16:00 & 937.0\\
3 February 2000 & Thursday & 16:00 -- 16:30 & 767.2\\
3 February 2000 & Thursday & 16:30 -- 17:00 & 580.4\\

4 February 2000 & Friday & 16:00 -- 16:30 & -569.7\\
10 February 2000 & Thursday & 14:30 -- 15:00 & -611.4\\
10 February 2000 & Thursday & 15:00 -- 15:30 & -658.9\\
10 February 2000 & Thursday & 15:30 -- 16:00 & -695.7\\
10 February 2000 & Thursday & 16:00 -- 16:30 & -728.5\\
\bottomrule
\end{tabular}
\end{table}

@@

Similarly, seasonal functional components can be obtained, using Fourier terms for example \autocite[reminiscent of][]{de2011forecasting}. This approach leads, with no additional effort, to handling seasonality with a fractional or varying period.

@@

@@

>>>>

\author{Alexander~Dokumentov, Rob J~Hyndman}
\addresses{\textbf{Alexander Dokumentov}\newline
Let's Forecast\newline Australia
\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University\newline Australia
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\newline Corresponding author\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Dokumentov, Hyndman: \@date}
\makeatother

\graphicspath{{figs/}}
\usepackage{amsmath,pdfpages}
\setlength{\arraycolsep}{3pt}
\setcounter{MaxMatrixCols}{20}
\newcommand{\argmin}{\operatorname{arg\,min}}
\newcommand{\sn}{\kappa}
\renewcommand{\vec}{\operatorname{vec}}
\newcommand{\diag}{\operatorname{diag}}
\renewcommand{\mod}{\operatorname{mod}}
\usepackage[section]{placeins}
\makeatletter\let\ftype@table\ftype@figure\makeatother
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

The first attempts to decompose time series into various components can be dated to the 19th century, when \textcite{Poynting1884} proposed price averaging as a tool for eliminating trend and seasonal fluctuations. Later, his approach was extended by \textcite{hooker1901suspension}, \textcite{spencer1904graduation} and \textcite{anderson1914nochmals}. Further research in this direction included that by \textcite{copeland1915statistical}, who was the first to attempt to extract the seasonal component, until \textcite{macaulay1930smooth} proposed a method which gradually became ``classical''. The work of Macaulay led to the Census II method \autocite{shiskin1957electronic} which became popular amongst national statistics offices. The Census II method has continued to evolve, with additional features including robustness, calendar effects, covariates, ARIMA extensions, and extensive diagnostics. Widely used versions of this approach have been X-11 \autocite{shishkin1967x}, X-11-ARIMA \autocite{dagum1988x11arima,ladiray2001seasonal}, X-12-ARIMA \autocite{findley1998new} and X-13-ARIMA-SEATS \autocite{findley2005some,Dagum2016}.

@@

Several model-based methods for seasonal decomposition have been developed, including the TRAMO/SEATS procedure which was developed at the Bank of Spain \autocite{Gomez2001}, the TBATS model of \textcite{delivera2011forecasting}, and various structural time series model approaches \autocite{harvey1990forecasting,Commandeur2011}. One big advantage of using a model for seasonal decomposition and adjustment is that it provides a natural way to compute confidence and prediction intervals.

@@

Note that the model subsumes the case of a linear trend, obtained when \(\sigma_L^2=0\).

@@

Again, by setting the smoothing parameters to zero, some simple special cases are obtained. Setting \(\sigma_i\) to zero eliminates the \(i\)th seasonal component (as it will be equal to zero); setting the variance in the time direction to zero causes the seasonal component to be periodic; setting the variance in the time-season direction to zero causes ????.

@@

\hypertarget{simulations}{%
\section{Simulations}\label{simulations}}

To demonstrate that our method estimates the appropriate time series components, we use simulations where we know the true underlying components. We will compare our results against two other decomposition methods that handle multiple seasonal components, STL and TBATS. STL \autocite{cleveland1990stl} is a well known decomposition method, and therefore a natural choice to use as a method to compare. The original methodology and implementation was designed for only one seasonal period. However, the implementation in the forecast package for R \autocite{Rforecast} will compute an STL decomposition iteratively with multiple seasonal periods. TBATS \autocite{delivera2011forecasting} is a forecasting method designed for multiple seasonalities; it provides a decomposition of the time series into components, and so can provide an alternative comparison.

Our simulated daily data consists of four components:
\begin{equation}
  \label{STR:many_parts2}
  y_{t} = T_{t} + \alpha S^{W}_{t} + \beta S^{Y}_{t} + \gamma R_{t} , \quad t=1,\dots,n,
\end{equation}
where \(T_{t}\) is a trend, \(S^{W}_{t}\) is a weekly seasonal component, \(S^{Y}_{t}\) is a yearly seasonal component, \(R_{t}\) is the remainder, and \(\alpha\), \(\beta\), and \(\gamma\) are parameters which control the contribution of the components to \(y_{t}\).

We use two data generating processes (DGPs) to simulate data. For the first DGP, we use a quadratic trend function with random coefficients, such that \(T_t^* = N_1(t + n/2(N_2-1))^2\) where \(N_1\) and \(N_2\) are independent N(0,1) random variables. Then \(T_t^*\) is normalized to give \(T_t\) with mean zero and unit variance. The seasonal components consist of five pairs of Fourier terms with random N(0,1) coefficients, which are then also normalized. Because this DGP has deterministic components, we refer to it as the ``Deterministic'' GDP.

For the second ``Stochastic'' DGP, the trend is given by an ARIMA(0,2,0) model with standard normal errors, which is then normalized. The seasonal components are also smooth and obtained using a similar process constrained to be periodic. Specifically, for each seasonal component we start with a vector of iid N(0,1) values of the length of one season, normalized and replicated to be of length \(n\). This is then integrated and normalized twice to make it smooth. So each seasonal component is as smooth as the trend, is centred on zero and is periodic.

Figure \ref{fig:examples} shows examples of simulated time series from each of the two DGPs with \(\alpha=\beta=1\) and \(\gamma=0.25\). We generated 1096 days in each case (one day more than three non-leap years of daily data).

\begin{figure}
\centering
\includegraphics{str_ijds_files/figure-latex/examples-1.pdf}
\caption{\label{fig:examples}Examples of time series generated by the two DGPs.}
\end{figure}

For comparison of the three decomposition methods, we use parameter sets \((\alpha, \beta, \gamma) = \{(1, 1, 0.2), (1, 1, 0.4), (1, 1, 0.6)\}\), and create 20 time series for each DGP and parameter set. We apply STR, STL and TBATS to each series and compute RMSE for each component (averaging across all data sets and days). We measure the significance of the differences between the methods using a linear model applied to the squared errors. The results are shown in Table \ref{tab:cbetable}. It is clear that STR outperforms STL in estimating all components other than the trend, and that STR outperforms TBATS in estimating all components other than the weekly seasonal pattern. In no case do the other methods outperform STR.

\begin{table}[!ht]

\caption{\label{tab:cbetable}RMSE over 100 simulations for each DGP and value of $\gamma$. Bold values indicate results that are significantly different from the STR values.}
\centering
\begin{tabular}[t]{rlrrrr}
\toprule
$\gamma$ & Method & Trend RMSE & Weekly RMSE & Yearly RMSE & Remainder RMSE\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Stochastic DGP}}\\
\hspace{1em}0.2 & STL & \textbf{0.0735} & \textbf{0.0564} & \textbf{0.1272} & \textbf{0.1495}\\
\hspace{1em}0.2 & TBATS & \textbf{0.1686} & \textbf{0.0169} & \textbf{0.1598} & \textbf{0.0579}\\
\hspace{1em}0.2 & STR & 0.0176 & 0.0150 & 0.0510 & 0.0544\\
\midrule
\hspace{1em}0.4 & STL & \textbf{0.0798} & \textbf{0.1149} & \textbf{0.2295} & \textbf{0.2653}\\
\hspace{1em}0.4 & TBATS & \textbf{0.3172} & \textbf{0.0344} & \textbf{0.3044} & \textbf{0.1182}\\
\hspace{1em}0.4 & STR & 0.0315 & 0.0306 & 0.0840 & 0.0929\\
\midrule
\hspace{1em}0.6 & STL & \textbf{0.0872} & \textbf{0.1713} & \textbf{0.3385} & \textbf{0.3887}\\
\hspace{1em}0.6 & TBATS & \textbf{0.5194} & \textbf{0.0516} & \textbf{0.4929} & \textbf{0.1852}\\
\hspace{1em}0.6 & STR & 0.0505 & 0.0447 & 0.1347 & 0.1470\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Deterministic GDP}}\\
\hspace{1em}0.2 & STL & \textbf{0.1279} & \textbf{0.0576} & \textbf{0.1387} & \textbf{0.1760}\\
\hspace{1em}0.2 & TBATS & \textbf{0.1849} & \textbf{0.0170} & \textbf{0.1803} & \textbf{0.0675}\\
\hspace{1em}0.2 & STR & 0.0552 & 0.0143 & 0.0700 & 0.0535\\
\midrule
\hspace{1em}0.4 & STL & \textbf{0.1129} & \textbf{0.1147} & \textbf{0.2366} & \textbf{0.2763}\\
\hspace{1em}0.4 & TBATS & \textbf{0.3232} & \textbf{0.0338} & \textbf{0.3146} & \textbf{0.1157}\\
\hspace{1em}0.4 & STR & 0.0541 & 0.0311 & 0.0930 & 0.0945\\
\midrule
\hspace{1em}0.6 & STL & \textbf{0.1362} & \textbf{0.1716} & \textbf{0.3428} & \textbf{0.3995}\\
\hspace{1em}0.6 & TBATS & \textbf{0.4834} & \textbf{0.0509} & \textbf{0.4698} & \textbf{0.1692}\\
\hspace{1em}0.6 & STR & 0.0793 & 0.0426 & 0.1304 & 0.1329\\
\bottomrule
\end{tabular}
\end{table}

@@

We obtained similar results using other decomposition methods applied to the same data including STL \autocite{cleveland1990stl}, TBATS \autocite{delivera2011forecasting} and X-13-ARIMA-SEATS \autocite{findley2005some,Dagum2016}. These comparisons are provided in the appendix.

@@

This data set has two seasonal patterns. The first pattern is a weekly seasonal pattern (\(m_1=48\times7=336\)) that represents the specific demand features that are attributable to a particular day of the week. The second pattern is a daily seasonal pattern (\(m_2=48\)) with the topology in Figure \ref{fig:topology_pdf}, which allows the model to distinguish between working days and holidays/weekends, and to make transitions between them. The pattern reflects the tendency to have a higher electricity demand during standard working hours and a lower demand at night. It also reflects the tendency to have different demand patterns on working days and holidays/weekends. A longer series would also have an annual seasonal pattern, but with only 115 days, this cannot be distinguished from the trend component.

@@

\begin{table}[!bh]

\caption{\label{tab:outlierstable}The ten residuals that are largest in absolute value after an STR decomposition.}
\centering
\begin{tabular}[t]{rlcr}
\toprule
Date & Day of week & Time period & Residual\\
\midrule
3 February 2000 & Thursday & 14:30 -- 15:00 & 687.3\\
10
3 February 2000 & Thursday & 15:00 -- 15:30 & 871.3\\
10
3 February 2000 & Thursday & 15:30 -- 16:00 & 937.0\\
10
3 February 2000 & Thursday & 16:00 -- 16:30 & 767.2\\
10
3 February 2000 & Thursday & 16:30 -- 17:00 & 580.4\\
10
4 February 2000 & Friday & 16:00 -- 16:30 & -569.7\\
10
10 February 2000 & Thursday & 14:30 -- 15:00 & -611.4\\
10
10 February 2000 & Thursday & 15:00 -- 15:30 & -658.9\\
10
10 February 2000 & Thursday & 15:30 -- 16:00 & -695.7\\
10
10 February 2000 & Thursday & 16:00 -- 16:30 & -728.5\\
\bottomrule
\end{tabular}
\end{table}

@@

Similarly, seasonal functional components can be obtained, using Fourier terms for example \autocite[reminiscent of][]{delivera2011forecasting}. This approach leads, with no additional effort, to handling seasonality with a fractional or varying period.

@@

\hypertarget{feature-engineering}{%
\subsection{Feature engineering}\label{feature-engineering}}

The STR model provides a way to identify new features in time series data, such as weekly or daily patterns. These can then be used within other algorithms. For example, they could be used to classify time series \autocite{catch22}, to find anomalous time series (as distinct from anomalous observations) \autocite{cikm2015}, to select a forecasting model \autocite{fforms}, or to create a weighted forecast combination \autocite{fforma}.

@@

\includepdf[pages=-]{str_ijds_appendix.pdf}
